A Generic and Efficient E-field Parallel Imaging Correlator for Next-Generation Radio Telescopes

### 2015-10-27 imaging_demo/ms.tex
Add new material in section 5, update figures
From 88cd7fb (master):
=====================

In ms.tex,

(*) Restructure section 5 into two subsections, one of which is a new
subsection on data throughput with the two algorithms
(*) Change figure 12
(*) Update figure 10
(*) Rearrange subsections in section 4
(*) Update figure 5
(*) Incorporate other changes suggested by Judd and Adam
(*) Minor changes to abstract, introduction and conclusions

% Please keep new commands to a minimum, and use \newcommand not \def to avoid
% overwriting existing commands. Example:
%\newcommand{\pcm}{\,cm$^{-2}$}	% per cm-squared
\newcommand{\Nant}{N_{\text{ant}}} 
\newcommand{\Npix}{N_{\text{pix}}}
\newcommand{\dif}{\mathrm{d}}

@@

% Abstract of the paper
\begin{abstract}
Modern radio telescopes are favouring densely packed array layouts 
consisting of a large number of antennas 
($N_\textrm{a}\gtrsim 1000$--$\gtrsim 10000$). Since the cost of traditional 
correlators scales as $\mathcal{O}(N_\textrm{a}^2)$, there will be a steep cost for 
realizing the full imaging potential of these powerful instruments. Through 
our generic and efficient E-field Parallel Imaging Correlator (EPIC), we present 
the first software demonstration of a generalized direct imaging algorithm 
known as the Modular Optimal Frequency Fourier (MOFF) imager. It takes 
advantage of the multiplication-convolution theorem of Fourier transforms. Not 
only does it bring down the cost to $\mathcal{O}(N_\textrm{a}\log N_\textrm{a})$ 
but can also image from irregularly arranged heterogeneous antenna arrays. 
EPIC is highly modular and parallelizable. It is implemented in object oriented 
Python and is publicly available. We have verified the images produced to be 
mathematically identical to those produced using traditional techniques. We have 
also validated our implementation on data observed with the Long Wavelength 
Array. Antenna systems with a dense filling factor consisting of a large number 
of antennas such as LWA, SKA, and HERA will gain a significant advantage by 
deploying EPIC. Inherent availability of calibrated time-domain images on 
timescales roughly equal to the writeout timescale of the digitizer will make 
it a prime candidate for transient searches of primarily Fast Radio Bursts (FRB) 
as well as slower planteary and exoplantenary phenomena. 
\end{abstract}

@@

Radio astronomy is entering an era in which interferometers of hundreds to
thousands of individual antennas are needed to achieve desired survey speeds.
Nowhere is this more apparent than at radio frequencies below 1.4 GHz. The study
of the history of hydrogen gas throughout the universe's evolution is pushing
technology development towards arrays of low-cost antennas with large fields of
view and densely packed apertures. Similarly, the search for transient objects
and regular monitoring of the time-dependent sky is driving instruments in the
same direction. A number of new telescopes are under development around the world
based on this new paradigm, including the Murchison Widefield Array (MWA;
\citealt{tin13,bow13}), the Precision Array for Probing the Epoch of Reionization
(PAPER; \citealt{par10}), the Hydrogen Epoch of Reionization 
Array\footnote{http://reionization.org} (HERA), the LOw Frequency ARray (LOFAR;
\citealt{van13}), the Canadian Hydrogen Intensity Mapping Experiment (CHIME,
\citealt{ban14}), the Long Wavelength Array (LWA, \citealt{ell13}), and the low
frequency component of the Square Kilometer Array Low Frequency Aperture Array
(SKA-Low \citealt{mel13}).

@@

For certain classes of radio arrays there is an alternative to the FX correlator
that can lower the computational burden by directly performing a spatial fast
Fourier transform (FFT) on the electric fields measured by each antenna in the
array at each time step, removing the cross-correlation step. This relieves the
computational scaling from the harsh $\Nant^2$ to the more gentle envelope of
$\Npix\log\Npix$, where $\Npix$ is the number of pixels in the Fourier transform
\citep[e.g.][]{mor11,teg09,teg10}. This architecture is often referred to as a
``direct imaging" correlator because it eliminates the intermediary cross-
correlation data products of the FX and earlier lag correlators, but instead
directly forms images from the electric field measurements.

Direct imaging correlators have begun to be explored on deployed arrays including
the Basic Element for SKA Training II (BEST-2) array \citep{fos14}, the Omniscope
\citep{zhe14}, and an earlier incarnation at higher frequencies with the intent
of pulsar timing \citep{oto94, dai00}. However, each of these examples make
assumptions about the redundancy of the array layout, and require the collecting
elements are identical. On the other hand, the MOFF algorithm achieves the same
$\Npix \log \Npix$ computational scaling without placing any restriction on
antenna placement, can accommodate non-identical beam patterns, and is 
a provably optimal mapping \citep{mor11}. This algorithm uses the antenna beam 
patterns to grid the electric field measurements to a regular grid in the software
holography/A-transpose fashion \citep{mor09,bha08,teg97b} before performing the
spatial FFT. This process has been shown to theoretically produce a data product
identical to images produced from the traditional FX correlator.

Here we present the first software implementation of the MOFF correlator, and
announce the public release of the E-field Parallel Imaging Correlator (EPIC)
code. We begin with a technical description of the algorithm in \S
\ref{sec:math}, then discuss our particular implementation in \S
\ref{sec:software}. We then verify the output data quality from our code in \S
\ref{sec:verify} by presenting simulated images from both the EPIC correlator
and comparing to a simulated FX correlator. We also demonstrate the performance
with real-world data from the LWA. In \S \ref{sec:analysis} we analyze the
scaling relationships of the algorithm. We identify specific array design classes
where the EPIC correlator is computationally more efficient than the FX
algorithm. We conclude and discuss future research prospects in \S
\ref{sec:conclusions}.

% Motivate MOFF from a technical and scientific standpoint. Refer to \citet{del07,mor09,mor11}.

@@

From here we adopt the matrix notation of \citet{mor11}, where vectors are 
represented with single coordinates, and matrices are represented by two 
coordinates denoting the spaces the operator transforms between. 
In this notation, the above measurement equation can be expressed as:
\begin{align}
  \mathbf{m}(\mathbf{v}) &= \widetilde{\mathbf{B}}(\mathbf{v},\mathbf{u})\,\mathbf{F}(\mathbf{u},\hat{\mathbf{s}})\,\mathbf{I}(\hat{\mathbf{s}}) + \mathbf{n}(\mathbf{v}),
\end{align}
where the sky brightness $\mathbf{I}(\hat{\mathbf{s}})$ is Fourier transformed using
$\mathbf{F}(\mathbf{u},\hat{\mathbf{s}})$ and the resultant spatial coherence
function is weighted and summed using the antenna power response,
$\widetilde{\mathbf{B}}(\mathbf{v},\mathbf{u})$ in $uv$-space sampled at the baseline
location to obtain the measured visibilities:
\begin{align}
  \mathbf{m}(\mathbf{v}) &= \left\langle \widetilde{\mathbf{E}}(\mathbf{a})\,\widetilde{\mathbf{E}}^\star(\mathbf{a}^\prime)\right\rangle_t, \label{eqn:matrix-cc-vis}
\end{align}
where $\mathbf{m}(\mathbf{v})$ denotes visibilities measured by cross-correlating
measured antenna electric fields over all possible pairs of $\mathbf{a}$ and
$\mathbf{a}^\prime$. It is the same as equation~\ref{eqn:cc-vis} written in
matrix notation.

@@

There are some important differences between the two techniques:
\begin{enumerate}
\item The time-averaging cannot be performed on a stochastic measurement but
  only on its statistical properties. In FX imaging, the visibilities measured
  between antenna pairs represent spatial correlations which can be time-averaged
  followed by gridding and imaging. However, in MOFF imaging both antenna and 
  gridded electric fields are stochastic and therefore must be imaged and squared 
  before time-averaging. 
\item In FX imaging, electric fields measured by antennas are not correlated with
  themselves and hence lack zero spacing measurements. In contrast, in MOFF
  imaging, since the gridded electric fields are imaged and squared, they retain 
  information from auto-correlated electric fields at zero spacing and thus yield 
  the true total power of the imaged field.
\end{enumerate}

@@

We have implemented the MOFF imaging technique in our ``E-field Parallel Imaging
Correlator'' -- a highly parallelized Object Oriented Python 
package,\footnote{EPIC package can be accessed at https://github.com/nithyanandan/EPIC}
now publicly available. Besides implementing the MOFF imaging algorithm it also
includes FX imaging using the software holography technique and a simulator for
generating electric fields from a sky model.

@@

Fig.~\ref{fig:FX-flowchart} shows the flowchart for software holographic
imaging from a FX correlator. The antenna-based F-engine is identical to that in
the MOFF processing. The electric field spectra from each antenna are then cross-
multiplied in the X-engine with those from all other antennas to obtain the
visibilities $V_\textrm{ij}(f)$ ($\mathbf{m}(\mathbf{v})$ in matrix notation).
They are calibrated and time-averaged to obtain $\langle V_\textrm{ij}(f)\rangle$
which are then gridded and imaged to obtain the image $I(f)$. The $I(f)$ obtained
from both techniques are identical as explained in \S\ref{sec:math}.
\begin{figure}
  \includegraphics[width=\columnwidth]{FX_flowchart.eps}
  \caption{A flowchart of FX imaging in EPIC. The FX process flow shares the
    F-engine with the MOFF process. Following the F-engine, the electric fields
    pass through the X-engine to obtain visibilities $V_\textrm{ij}(f)$ which are
    calibrated and time-averaged. Then they are gridded to obtain the gridded
    visibilities $V_\textrm{g}(f)$ which are then imaged to obtain the image
    $I(f)$.}
  \label{fig:FX-flowchart}
\end{figure}

@@

\par\medskip
\noindent {\bf Temporal Fourier transform}
\par\medskip
\noindent This module is common to the MOFF and FX imaging techniques. Time 
samples of electric fields measured by the antenna and digitized by the A/D 
converter is Fourier transformed to generate electric field spectra. This step 
can be parallelized by antennas as shown in Fig.~\ref{fig:f-engine}. The output 
is then fed to either MOFF and FX imaging pipelines.

@@

In the most general case, this gridding kernel could contain information on the
$w$-projection effect, and even other time-dependent ionospheric effects. For a
stationary antenna array in the absence of any time-dependent effects, this
mapping must only be determined once in the antenna array coordinate frame. The
antenna-to-grid mapping matrix, $\mathbf{M}(\mathbf{r},\mathbf{a})$ is described
as a transformation matrix from the space of measured electric fields by the 
antennas ($\mathbf{a}$) to the antenna array grid denoted by the coordinate 
$\mathbf{r}$. Since each antenna occupies a footprint typically the size of its 
aperture, $\mathbf{M}(\mathbf{r},\mathbf{a})$, which is generally of size
$N_\textrm{grid}\times N_\textrm{a}$, reduces to a sparse block-diagonal matrix
with only $N_\textrm{a}$ blocks and roughly $N_\textrm{k}$ non-zero entries per
block. This sparse matrix is stored in a Compressed Sparse Row (CSR) format. 
Fig.~\ref{fig:a2g-mapping} illustrates the antenna-to-grid mapping
matrix and the grid containing the mapped aperture footprints of the antennas.

\begin{figure}
  \includegraphics[width=\columnwidth]{a2g_mapping.eps}
  \caption{Block diagram of an antenna-to-grid mapping. A sparse block-diagonal
    matrix of total size $N_\textrm{grid}\times N_\textrm{a}$ is created where each
    block contains roughly the number of pixels covered by the respective kernel.
    The antenna aperture illumination kernels do not have to be identical to each
    other. A discrete set of arbitrarily placed antennas are now placed onto a
    regular grid.}
  \label{fig:a2g-mapping}
\end{figure}

@@

In MOFF imaging, these are spatially Fourier transformed followed by a squaring
operation at every timestamp for every frequency channel. In FX imaging, the
spatial Fourier transform is performed only once per integration timescale and
does not include a squaring operation.

@@

In contrast, visibilities measured by an antenna are statistically stable within
an integration time interval. Hence, they are averaged after calibration as shown
in equation~\ref{eqn:cc-vis}. It is advantageous to average them in visibilities
before imaging because the repeated cost of spatial FFT can be avoided. Since 
this averaging has been performed already on the visibilities over an integration 
timescale, the imaging step has to be performed only once per integration cycle. 
% FX imaging holds this advantage as long as the square of the number of antennas 
% is smaller than the number of pixels on the image.

@@

\begin{figure*}
  \includegraphics[width=\linewidth]{EPIC-modules.eps}
  \caption{Software architecture of EPIC with core modules, their essential
    attributes and functions. The Antenna module forms the fundamental building 
    block. It consists of electric field time-series and spectra and the 
    F-engine that performs a temporal FFT to obtain electric field spectra from 
    the time-series. The Interferometer module is made of a pair of Antenna 
    modules. Its main function is the X-engine (FX or XF) to produce visibility 
    spectra. The Antenna Array module is made of all individual Antenna modules
    as its components and contains collective properties about the antenna 
    subsystems. Its core function is the creation of antenna-to-grid mapping and 
    that of gridded aperture weights and electric fields. The Interferometer 
    Array module is very similar in principle to the Antenna Array module 
    except it operates on cross-correlations and produces gridded visibilities.
    The Image module takes gridded electric fields or visibilities and performs
    a two-dimensional spatial FFT (and squares the intermediate image in case 
    of the former) to produce output images.}
  \label{fig:software-modules}
\end{figure*}

\subsubsection{Antenna module}

The Antenna module is a fundamental building block upon which all the other 
modules are built. There is one Antenna module per antenna each having 
attributes -- the propagated electric field time-series, $E(t)$, and spectrum 
$E(f)$ for both polarizations. The most important function inside this module 
is the F-engine that Fourier transforms time-series electric field data into 
spectra.

The other function (not shown in the figure) is to update the data as new data 
keeps streaming in. This can also be parallelized. Another important attribute
(not shown in figure) are antenna flags for each polarization appropriate for 
the data stream being held by the Antenna module.

\subsubsection{Interferometer module}

The Interferometer module holds the attributes and functions pertaining to
a pair of antennas and represents the cross-correlation information obtained
from the pair. Its important attributes are the two Antenna modules. It also
contains four cross-polarized visibility time-series and spectra.

The most important function of the Interferometer module is the X-engine. 
This is essentially a software analog of hardware correlators of real 
telescope systems. The X-engine can be toggled between two states of operation,
namely, the FX and XF modes. The FX mode obtains the electric field spectra, 
$E(f)$ from the individual Antenna modules inside the Interferometer module and
multiplies the two to obtain visibility spectra, $V(f)$. On the other hand, the
XF mode cross-correlates the electric field time-series from its Antenna modules
to obtain the visibilities as a function of lags, $V_t(t)$, which is then Fourier 
transformed to obtain $V(f)$. Both modules can operate on dual-polarizations to
obtain all four cross-polarizations.

The other attributes (not shown in figure) are the flags applicable for each
cross-polarization for the current data stream. Similar to the Antenna module,
it has an update function that can update the visibilities $V_t(t)$ or $V(f)$
directly rather than through the electric fields of its component antennas. 
This functionality is to allow EPIC to operate while attached to the backend 
of traditional correlator systems. Currently, this feature is not utilized 
for purposes of this paper.

The Interferometer module forms the fundamental unit for the InterferometerArray
module (to be discussed below) and in general for visibility based correlator 
and imaging systems.

\subsubsection{Antenna Array module}

The AntennaArray module consists of all the Antenna modules as its attributes
and represents the collective properties of its component antennas. By virtue of 
holding each antenna data independently in their respective modules the F-engine 
for the entire array can be distributed to the F-engines of the component 
antenna modules thus achieving a highly parallelized F-engine while emulating 
real telescope systems.

The most important attributes held by this module are the antenna aperture 
illumination weights and electric fields projected on the grid using the 
gridding convolution method described above and implemented by the gridding 
function (denoted by grid()) in this module. Significant parts of the 
antenna-to-grid mapping and gridding convolution are parallelizable 
across antennas and frequencies.

@@

\subsubsection{Interferometer Array module}

Similar to the AntennaArray module, the InterferometerArray module consists of
individual Interferometer modules. It can parallelize the correlator operations
by distributing the X-operation over the X-engines of its component 
Interferometer modules. The interferometer-to-grid mapping and gridding 
convolution are very similar in nature to that of the Antenna Array module. 
Flag-based grid weights, stacking and ability to handle all four 
cross-polarizations are built into this module.

\subsubsection{Image module}

The Image module is built as a general purpose module that can switch between 
operating on gridded electric fields or visibilities. At its heart, it consists
of a two-dimensional FFT where the padding can be specified by the user to 
control the resolution in the output images. In case of electric field based
imaging, there is an additional step of squaring the holographic electric field
images. This is packaged into the function annotated as imagr().

@@

In order to prove the accuracy of the EPIC code, we verify the images produced 
through simulations. We simulate electric field streams from a model sky and 
process the data through both the MOFF and visibility based imaging algorithms. 
We then compare the output images to demonstrate their equivalence.

@@

Before diving into the details of simulations and comparison with visibility 
based imaging, we describe the elimination of a minor mathematical difference 
between the two techniques. The squaring operation under MOFF imaging in the 
image plane introduces antenna auto-correlations around the zero spacing in 
the $uv$-plane which are absent in traditional visibility-based imaging. In 
order to facilitate a robust comparison, these auto-correlations are removed 
from both the $uv$ and image planes, which is otherwise not an essential part 
of the core algorithm. We describe below how they can be removed from the $uv$ 
plane in a straightforward manner.

@@

We inverse Fourier transform the squared images and beams back to the $uv$ 
plane and subtract the estimated auto-correlation kernel scaled to the peak 
value centered at the zero spacing pixel. The final averaged image is obtained 
by Fourier transforming the $uv$ plane data and weights with the 
auto-correlations subtracted to the image plane. These images are now comparable 
to those obtained from visibility-based imaging. This step of removing 
auto-correlations is optional and required to be performed only once per 
integration timescale and does not add significant cost to the full operation.

Fig.~\ref{fig:MOFF-FX-image} shows the dirty images (top) and synthesized beams
(bottom) obtained with antenna-based MOFF and FX visibility-based imaging 
algorithms packaged in EPIC. The antenna auto-correlations that correspond to 
zero spacing have been removed from the MOFF image and the corresponding 
synthesized beam. The sky positions of the simulated sources are indicated by 
solid black circles. The reconstructed sky image has the simulated sources at the
expected sky positions in either case. Both algorithms result in images and 
synthesized beams that are well matched with each other. Their fluxes are 
modulated by a multiplicative power pattern corresponding to that of a uniform
square aperture.

@@

>>>>

% Please keep new commands to a minimum, and use \newcommand not \def to avoid
% overwriting existing commands. Example:
%\newcommand{\pcm}{\,cm$^{-2}$}	% per cm-squared
\newcommand{\Nant}{N_\textrm{a}}
\newcommand{\Ngrid}{N_\textrm{g}} 
\newcommand{\Npix}{N_{\text{pix}}}
\newcommand{\dif}{\mathrm{d}}

@@

% Abstract of the paper
\begin{abstract}
Modern radio telescopes are favouring densely packed array layouts 
consisting of a large number of antennas 
($\Nant\gtrsim 1000$--$\gtrsim 10000$). Since the complexity of traditional 
correlators scales as $\mathcal{O}(\Nant^2)$, there will be a steep cost for 
realizing the full imaging potential of these powerful instruments. Through 
our generic and efficient E-field Parallel Imaging Correlator (EPIC), we present 
the first software demonstration of a generalized direct imaging algorithm 
known as the Modular Optimal Frequency Fourier (MOFF) imager. It takes 
advantage of the multiplication-convolution theorem of Fourier transforms. Not 
only does it bring down the cost to $\mathcal{O}(\Nant\log \Nant)$ 
but can also image from irregularly arranged heterogeneous antenna arrays. 
EPIC is highly modular and parallelizable. It is implemented in object oriented 
Python and is publicly available. We have verified the images produced to be 
mathematically identical to those produced using traditional techniques. We have 
also validated our implementation on data observed with the Long Wavelength 
Array. Antenna systems with a dense filling factor consisting of a large number 
of antennas such as LWA, SKA, and HERA will gain a significant advantage by 
deploying EPIC. Inherent availability of calibrated time-domain images on 
timescales roughly equal to the writeout timescale of the digitizer and vastly 
lower bandwidth relative to visibility based systems will make it a prime 
candidate for transient searches of Fast Radio Bursts (FRB) as well as 
planetary and exoplantenary phenomena. 
\end{abstract}

@@

Radio astronomy is entering an era in which interferometers of hundreds to
thousands of individual antennas are needed to achieve desired survey speeds.
Nowhere is this more apparent than at radio frequencies below 1.4 GHz. The study
of the history of hydrogen gas throughout the universe's evolution is pushing
technology development towards arrays of low-cost antennas with large fields of
view and densely packed apertures. Similarly, the search for transient objects
and regular monitoring of the time-dependent sky is driving instruments in the
same direction with the added requirement of fast read-outs. A number of new 
telescopes are under development around the world based on this new paradigm, 
including the Murchison Widefield Array (MWA;\citealt{tin13,bow13}), the 
Precision Array for Probing the Epoch of Reionization (PAPER; \citealt{par10}), 
the Hydrogen Epoch of Reionization Array\footnote{http://reionization.org} 
(HERA), the LOw Frequency ARray (LOFAR;\citealt{van13}), the Canadian Hydrogen 
Intensity Mapping Experiment (CHIME,\citealt{ban14}), the Long Wavelength 
Array (LWA, \citealt{ell13}), and the low frequency component of the Square 
Kilometer Array Low Frequency Aperture Array (SKA1-Low \citealt{mel13}).

@@

For certain classes of radio arrays there is an alternative to the FX correlator
that can lower the computational burden by directly performing a spatial fast
Fourier transform (FFT) on the electric fields measured by each antenna in the
array at each time step, removing the cross-correlation step. This relieves the
computational scaling from the harsh $\Nant^2$ to the more gentle envelope of
$\Ngrid\log\Ngrid$, where $\Ngrid$ is the number of grid pixels in the Fourier 
transform \citep[e.g.][]{mor11,teg09,teg10}. This architecture is often referred 
to as a ``direct imaging" correlator because it eliminates the intermediary 
cross-correlation data products of the FX and earlier lag correlators, but 
instead directly forms images from the electric field measurements.

Direct imaging correlators have begun to be explored on deployed arrays including
the Basic Element for SKA Training II (BEST-2) array \citep{fos14}, the Omniscope
\citep{zhe14}, and an earlier incarnation at higher frequencies with the intent
of pulsar timing \citep{oto94, dai00}. However, each of these examples make
assumptions about the redundancy of the array layout, and require the collecting
elements are identical. On the other hand, the MOFF algorithm achieves the same
$\Ngrid \log \Ngrid$ computational scaling without placing any restriction on
antenna placement, can accommodate non-identical beam patterns, and is 
a provably optimal mapping \citep{mor11}. This algorithm uses the antenna beam 
patterns to grid the electric field measurements to a regular grid in the software
holography/A-transpose fashion \citep{mor09,bha08,teg97b} before performing the
spatial FFT. This process has been shown to theoretically produce a data product
identical to images produced from the traditional FX correlator.

Here we present the first software implementation of the MOFF correlator, and
announce the public release of the E-field Parallel Imaging Correlator (EPIC)
code. We begin with a technical description of the algorithm in \S\ref{sec:math}, 
then discuss our particular implementation in \S\ref{sec:software}. We then 
verify the output data quality from our code in \S\ref{sec:verify} by presenting 
simulated images from both the EPIC correlator and comparing to a simulated FX 
correlator. We also demonstrate the performance with real-world data from the 
LWA. In \S \ref{sec:analysis} we analyze the scaling relationships of the 
algorithm. We identify specific array design classes where the EPIC correlator 
is computationally more efficient and in the field of transients, demands 
significantly lesser bandwidth relative to visibility based approaches. We 
conclude and discuss future research prospects in \S\ref{sec:conclusions}.

@@

Hereafter, we adopt the matrix notation of \citet{mor11}, where vectors are 
represented with single coordinates, and matrices are represented by two 
coordinates denoting the spaces the operator transforms between. 
In this notation, the above measurement equation can be expressed as:
\begin{align}
  \mathbf{m}(\mathbf{v}) &= \widetilde{\mathbf{B}}(\mathbf{v},\mathbf{u})\,\mathbf{F}(\mathbf{u},\hat{\mathbf{s}})\,\mathbf{I}(\hat{\mathbf{s}}) + \mathbf{n}(\mathbf{v}),
\end{align}
where the sky brightness $\mathbf{I}(\hat{\mathbf{s}})$ is Fourier transformed using
$\mathbf{F}(\mathbf{u},\hat{\mathbf{s}})$ and the resultant spatial coherence
function is weighted and summed using the antenna power response,
$\widetilde{\mathbf{B}}(\mathbf{v},\mathbf{u})$ in $uv$-space sampled at the baseline
location to obtain the measured visibilities:
\begin{align}
  \mathbf{m}(\mathbf{v}) &= \left\langle \widetilde{\mathbf{E}}(\mathbf{a})\,\widetilde{\mathbf{E}}^\star(\mathbf{a}^\prime)\right\rangle_t, \label{eqn:matrix-cc-vis}
\end{align}
where $\mathbf{m}(\mathbf{v})$ denotes visibilities measured by cross-correlating
measured antenna electric fields over all possible pairs of $\mathbf{a}$ and
$\mathbf{a}^\prime$. It is the same as equation~\ref{eqn:cc-vis} written in
matrix notation.

@@

There are some important differences between the two techniques:
\begin{enumerate}
\item The time-averaging cannot be performed on a stochastic measurement but
  only on its statistical properties. In visibility based imaging, the 
  visibilities measured between antenna pairs represent spatial correlations 
  which can be time-averaged followed by gridding and imaging. However, in MOFF 
  imaging both antenna and gridded electric fields are stochastic and therefore 
  must be imaged and squared before time-averaging. 
\item In visibility based imaging, electric fields measured by antennas are not 
  correlated with themselves and hence lack zero spacing measurements. In 
  contrast, in MOFF imaging, since the gridded electric fields are imaged and 
  squared, they retain information from auto-correlated electric fields at zero 
  spacing and thus yield the true total power of the imaged field.
\end{enumerate}

@@

We have implemented the MOFF imaging technique in our ``E-field Parallel Imaging
Correlator'' -- a highly parallelized Object Oriented Python 
package,\footnote{EPIC package can be accessed at https://github.com/nithyanandan/EPIC}
now publicly available. Besides implementing the MOFF imaging algorithm it also
includes visibility based imaging using the software holography technique and a 
simulator for generating electric fields from a sky model.

@@

Fig.~\ref{fig:FX-flowchart} shows the flowchart for a visibility based software 
holographic imaging from a FX correlator. The antenna-based F-engine is identical 
to that in the MOFF processing. The electric field spectra from each antenna are 
then cross-multiplied in the X-engine with those from all other antennas to 
obtain the visibilities $V_\textrm{ij}(f)$ ($\mathbf{m}(\mathbf{v})$ in matrix 
notation). They are calibrated and time-averaged to obtain 
$\langle V_\textrm{ij}(f)\rangle$ which are then gridded and imaged to obtain the 
image $I(f)$. The $I(f)$ obtained from both techniques are identical as explained 
in \S\ref{sec:math}.
\begin{figure}
  \includegraphics[width=\columnwidth]{FX_flowchart.eps}
  \caption{A flowchart of visibility based software holographic imaging in EPIC. 
    The FX process flow shares the F-engine with the MOFF process. Following the 
    F-engine, the electric fields pass through the X-engine to obtain 
    visibilities $V_\textrm{ij}(f)$ which are calibrated and time-averaged. Then 
    they are gridded to obtain the gridded visibilities $V_\textrm{g}(f)$ which 
    are then Fourier transformed to obtain the image.
    $I(f)$.}
  \label{fig:FX-flowchart}
\end{figure}

@@

\par\medskip
\noindent {\bf Temporal Fourier transform}
\par\medskip
\noindent This module is common to the MOFF and visibility based imaging 
techniques. Time samples of electric fields measured by the antenna and 
digitized by the A/D converter is Fourier transformed to generate electric 
field spectra. This step can be parallelized by antennas as shown in 
Fig.~\ref{fig:f-engine}. The output is then fed to either MOFF or visibility 
based imaging pipelines.

@@

In the most general case, this gridding kernel could contain information on the
$w$-projection effect, and even other time-dependent ionospheric effects. For a
stationary antenna array in the absence of any time-dependent effects, this
mapping must only be determined once in the antenna array coordinate frame. The
antenna-to-grid mapping matrix, $\mathbf{M}(\mathbf{r},\mathbf{a})$ is described
as a transformation matrix from the space of measured electric fields by the 
antennas ($\mathbf{a}$) to the antenna array grid denoted by the coordinate 
$\mathbf{r}$. Since each antenna occupies a footprint typically the size of its 
aperture, $\mathbf{M}(\mathbf{r},\mathbf{a})$, which is generally of size
$\Ngrid\times \Nant$, reduces to a sparse block-diagonal matrix
with only $\Nant$ blocks and roughly $N_\textrm{k}$ non-zero entries per
block. This sparse matrix is stored in a Compressed Sparse Row (CSR) format. 
Fig.~\ref{fig:a2g-mapping} illustrates the antenna-to-grid mapping
matrix and the grid containing the mapped aperture footprints of the antennas.

\begin{figure}
  \includegraphics[width=\columnwidth]{a2g_mapping.eps}
  \caption{Block diagram of an antenna-to-grid mapping. A sparse block-diagonal
    matrix of total size $\Ngrid\times \Nant$ is created where each
    block contains roughly the number of pixels covered by the respective kernel.
    The antenna aperture illumination kernels do not have to be identical to each
    other. A discrete set of arbitrarily placed antennas are now placed onto a
    regular grid.}
  \label{fig:a2g-mapping}
\end{figure}

@@

In MOFF imaging, these are spatially Fourier transformed followed by a squaring
operation at every timestamp for every frequency channel. In visibility based 
imaging, the spatial Fourier transform is performed only once per integration 
timescale and does not include a squaring operation.

@@

In contrast, visibilities measured by an antenna are statistically stable within
an integration time interval. Hence, they are averaged after calibration as shown
in equation~\ref{eqn:cc-vis}. It is advantageous to average them in visibilities
before imaging because the repeated cost of spatial FFT can be avoided. Since 
this averaging has been performed already on the visibilities over an integration 
timescale, the imaging step has to be performed only once per integration cycle.

@@

\begin{figure*}
  \includegraphics[width=\linewidth]{EPIC-modules.eps}
  \caption{Software architecture of EPIC with core modules, their essential
    attributes and functions. The antenna module forms the fundamental building 
    block. It consists of electric field time-series and spectra and the 
    F-engine that performs a temporal FFT to obtain electric field spectra from 
    the time-series. The interferometer module is made of a pair of antenna 
    modules. Its main function is the X-engine (FX or XF) to produce visibility 
    spectra. The antenna array module is made of all individual antenna modules
    as its components and contains collective properties about the antenna 
    subsystems. Its core function is the creation of antenna-to-grid mapping and 
    that of gridded aperture weights and electric fields. The interferometer 
    array module is very similar in principle to the antenna array module 
    except it operates on cross-correlations and produces gridded visibilities.
    The image module takes gridded electric fields or visibilities and performs
    a two-dimensional spatial FFT (and squares the intermediate image in case 
    of the former) to produce output images. Broadly, the MOFF algorithm is 
    implemented by modules below the horizontal dashed line while the visibility 
    based imaging uses modules above the line. The exact processing pathway 
    implementing the MOFF algorithm is shown in bolded modules.}
  \label{fig:software-modules}
\end{figure*}

\subsubsection{Antenna Module}

The antenna module is a fundamental building block upon which all the other 
modules are built. There is one antenna module per antenna each having 
attributes -- the propagated electric field time-series, $E(t)$, and spectrum 
$E(f)$ for both polarizations. The most important function inside this module 
is the F-engine that Fourier transforms time-series electric field data into 
spectra.

The other function (not shown in the figure) is to update the data as new data 
keeps streaming in. This can also be parallelized. Another important attribute
(not shown in figure) are antenna flags for each polarization appropriate for 
the data stream being held by the module.

\subsubsection{Interferometer Module}

The interferometer module holds the attributes and functions pertaining to
a pair of antennas and represents the cross-correlation information obtained
from the pair. Its primary attributes are the two antenna modules. It also
contains four cross-polarized visibility time-series and spectra.

The critical component of the interferometer module is the X-engine. This is 
essentially a software analog of hardware correlators of real telescope 
systems. The X-engine can be toggled between two states of operation, 
namely, the FX and XF modes. The FX mode obtains the electric field spectra, 
$E(f)$ from the individual antenna modules inside this module and multiplies 
the two to obtain visibility spectra, $V(f)$. On the other hand, the XF mode 
cross-correlates the electric field time-series from its Antenna modules
to obtain the visibilities as a function of lags, $V_t(t)$, which is then Fourier 
transformed to obtain $V(f)$. Both modules can operate on dual-polarizations to
obtain all four cross-polarizations.

The other attributes (not shown in figure) are the flags applicable for each
cross-polarization for the current data stream. Similar to the antenna module,
it has an update function that can update the visibilities $V_t(t)$ or $V(f)$
directly rather than through the electric fields of its component antennas. 
This functionality is to allow EPIC to operate while attached to the backend 
of traditional correlator systems. This feature is not utilized for purposes 
of this paper.

This module forms the fundamental unit for the interferometer array module 
(to be discussed below) and in general for visibility based correlator and 
imaging systems.

\subsubsection{Antenna Array Module}

The antenna array module consists of all the antenna modules as its attributes
and represents the collective properties of its component antennas. By virtue of 
holding each antenna data independently in their respective modules the F-engine 
for the entire array can be distributed to the F-engines of the component 
antenna modules thus achieving a highly parallelized F-engine while emulating 
real telescope systems.

The primary attributes held by this module are the antenna aperture 
illumination weights and electric fields projected on the grid using the 
gridding convolution method described above and implemented by the gridding 
function in this module. Significant parts of the antenna-to-grid mapping and 
gridding convolution are parallelizable across antennas and frequencies.

@@

\subsubsection{Interferometer Array Module}

Similar to the antenna array module, the interferometer array module consists of
individual interferometer modules. It can parallelize the correlator operations
by distributing the X-operation over the X-engines of its component 
interferometer modules. The interferometer-to-grid mapping and gridding 
convolution are very similar in nature to that of the antenna array module. 
Flag-based grid weights, stacking and ability to handle all four 
cross-polarizations are built into this module.

\subsubsection{Image Module}

The image module is built as a general purpose module that can switch between 
operating on gridded electric fields or visibilities. At its heart, it consists
of a two-dimensional spatial FFT where the padding can be specified by the user 
to control the resolution in the output images. In case of MOFF imaging, there 
is an additional step of squaring the holographic electric field images.

@@

In order to verify the accuracy of the EPIC code, we characterize the images 
produced through simulations. We simulate electric field streams from a model 
sky and process the data through both the MOFF and visibility based imaging 
algorithms. We then compare the output images to demonstrate their equivalence.

@@

Before the outputs can be compared, we describe the elimination of a minor 
mathematical difference between the two techniques. The squaring operation under 
MOFF imaging in the image plane introduces antenna auto-correlations around the 
zero spacing in the $uv$-plane which are absent in traditional visibility based 
imaging. In order to facilitate a robust comparison between MOFF and visibility 
based imaging techniques, these auto-correlations are removed from both the MOFF 
algorithm output, which is otherwise not an essential part of the core algorithm. 
We describe below how they are removed.

@@

We inverse Fourier transform the squared images and beams back to the $uv$ 
plane and subtract the estimated auto-correlation kernel scaled to the peak 
value centered at the zero spacing pixel. The final averaged image is obtained 
by Fourier transforming the $uv$ plane data and weights with the 
auto-correlations subtracted to the image plane. These images are now comparable 
to those obtained from visibility based imaging. This step of removing 
auto-correlations is optional and required to be performed only once per 
integration timescale and does not add significant cost to the full operation.

@@

\subsection{Processing Volumes}
